---
title: "Biost537_hw1"
author: "Qin Li"
date: "1/15/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
getwd()
library(survival)
library(flexsurv)
library(foreign)
source("fitparametric.R")
library(numDeriv)
library(tidyr)
library(knitr)
```

## Problem 1

### a). 
The individual A is left truncated at age 32 since A does not have the cancer at the time of enrollment, and A will also be right censored 30 years later when the study ends. 

### b). 
The patient B is left truncated at age 39, and will be interval censored since B develop the cancer between the fourth and fifth exam after enrollment. 

### c). 
The individual C is left truncated at age 55 for age at onset, and is right censored at age 61 for death from heart failure.

### d). 
The individual D is left truncated at age 45 for age at onset, and is right censored at age 56 for moving away and not diagnosed with breast cancer. 

If instead of age at onset of breast cancer we were interested in studying the time from enrollment into the study until onset of breast cancer, the left truncation is no longer valid, but the censoring information will remain the same for all individuals. 

## Problem 2

### a). 

The study will have right censored data, it might due to death to other disease, or moving to other places. It is also likely that the person develop to the stage that they may dead after the study period, or develop to the stage that they may need a surgery. 

Both left truncation will eliminate patients from the study, since they would take a longer time to need a surgery or death than other patients who may need a surgery or death. 

### b). 
I don't agree with the statement. Age might be an important factor in whether patient may have the diagnosis, it is likely that patient are more likely develop the disease at certain age, and may also more likely to have bowel resection or death. Therefore, A may also affected by selection bias. 


## Problem 3

### a). 

312 Mayo Clinic patients with primary biliary cirrhosis.

### b). 

Enrolled in either of two double-blind, placebo-controlled, randomized clinical trials at the Mayo Clinic. 

### c).

Death 

### d). 

10 years

### e). 

Transplantation, lost to follow up, end of study

### f). 

Liver transplantation might be related to the outcome of interest. Only patients at certain level of conditions will be allowed for liver transplantation, whether the condition need to be better than people who cannot do liver transplantation or worse, it is associated with the chance of death. 
It is not clear that are the causes of loss to follow up. 
Since the study is about 10 years, the censored patients due to end of study is not informative, therefore it should not be related to the outcome of interest. 

### g). 

For people at high risk, approximated 83% of individuals die within 5 years; for people at medium risk, approximated 43% of individuals die within 5 years; for people at low risk, approximated 10% of individuals die within 5 years. 

## Problem 4

### a). 

The average follow-up time is 402.57 days, and the proportions of censored observations is 0.37.


```{r Q4}
addicts <- read.csv("addicts.csv",header = TRUE)

avg.time <- mean(addicts$time)
prop.event <- 1- mean(addicts$event)

s.addicts <- with(addicts, Surv(time, event))
surv.addicts <- Surv(time =addicts$time, event = addicts$event, type = "right")
survfit.add <- survfit(surv.addicts~1, data = addicts, conf.type = "log-log")
mod.exp <- fitparametric(s.addicts,dist="exp")
mod.weibull <- fitparametric(s.addicts,dist="weibull")
mod.gengamma <- fitparametric(s.addicts,dist="gengamma")



```

### b). 

See the tables below


```{r Q42,echo=FALSE}
kable(cbind(mod.exp$coeff,mod.exp$loglik),caption = "exp coefs",
      col.names = c("estimate","ci.lower","ci.higher","se","loglike"))

kable(cbind(mod.weibull$coeff,mod.weibull$loglik), 
      caption = "weibull coefs",
      col.names = c("estimate","ci.lower","ci.higher","se","loglike"))

kable(cbind(mod.gengamma$coeff,mod.gengamma$loglik), caption = "gengamma coefs",
      col.names = c("estimate","ci.lower","ci.higher","se","loglike")) 
```

### c). 

The exponential model is constantly below the kaplan-meier curve until time is greater than 600 days, whereas weibull and generalized gamma models are similar and are generally better than the exponential model, because they do not differ much compare to the non-parametric estimators curve. 

```{r Q4c,echo=FALSE}
fitexp <- flexsurvreg(Surv(time, event)~1, data = addicts, dist = "exp")
fitweibull <- flexsurvreg(Surv(time, event)~1, data = addicts, dist = "weibull")
fitggamma <- flexsurvreg(Surv(time, event)~1, data = addicts, dist = "gengamma")

plot(survfit(Surv(time,event)~1, data = addicts),conf.int = FALSE,
     xlab = "time in days",ylab = "survival prob.")

lines(fitexp, col = 2, ci = FALSE,lty = 3)
lines(fitweibull, col = 3, ci = FALSE, lty = 3)
lines(fitggamma, col = 4, ci = FALSE, lty = 3)

legend("topright", legend = c("exp","weibull","gengamma","nonparametric est"), fill = c(2,3,4,1))

```
### d). 

The log likelihood ratio is 2*(-1114.36-(-1114.92)) = 1.12 on 1 degree of freedom, which is not significant (p=0.29), therefore we fail to reject the null  that the Weibull is an appropriate simplification of the generalized gamma model. 

```{r Q4d}
1-pchisq(1.12,1)

```
### e). 

i. The median time until exit from maintenance is 457.84 (95%CI: 396.99,518.49).
ii. The probability that no exit will occur by one year is 1.23 (95%CI: 1.06, 1.39). 
iii. The probability that no exit will occur by two years given that no exit has occured by one year is 0.49 (95%CI: 0.42,0.57). 

```{r Q4e,echo = F}
fitparametric(s.addicts, dist = "weibull", feature = "quantile")
fit_weibull1 <- fitparametric(s.addicts, dist = "weibull", feature = "survival", 
                              t = 365)
fit_weibull2 <- fitparametric(s.addicts, dist = "weibull",
                              feature ="condsurvival", t = 730, t0 = 365)
```
f). 

The 95% CI for p is (1.06,1.39) which does not contain 1, so Weibull is preferred. The log likelihood ratio is 2*(-1114.92-(-1118.93)) = 8.02 on 1 degree of freedom (p = 0.0046), so clearly Weibull is preferred.  

g). 

We have an estimate of $\delta$, i.e. 
$$\hat{\delta}=\hat{\lambda}_1 - \hat{\lambda}_2=0.002048423-0.0007723286= 0.001276094.$$
If we can derive the standard error of $\hat{\delta}$, $SE(\hat{\delta})$, we can construct a Wald statistic
$$\hat{T}_{W}=\frac{|\hat{\delta}-\delta_0|}{SE(\hat{\delta})},$$ 
where $\delta_0=0$ under the null hypothesis that the time to death distribution doesn't differ between those with or without diabetes. We have $\delta = 0.0013$, and calculated the two-sided wald test p value as 6.404186e-08. Thus, at $\alpha$ = 0.05 level, we reject the null that the distribution of tme to exit from maintenance differs significantly by clinic. 


```{r Q4g,echo=FALSE}
fit_exp_clinic1 <- flexsurvreg(Surv(time, event) ~ 1, 
                                  data = addicts[addicts$clinic ==1, ],
                                  dist = "exponential")
fit_exp_clinic2 <- flexsurvreg(Surv(time, event) ~ 1, 
                                  data = addicts[addicts$clinic ==2, ],
                                  dist = "exponential")

delta <- fit_exp_clinic1$res[1] - fit_exp_clinic2$res[1]
delta_se <- sqrt(fit_exp_clinic1$res[4] ^ 2 + 
                     fit_exp_clinic2$res[4] ^ 2)
T_W <- abs(delta) / delta_se
T_W

2 * pnorm(-T_W) 

```

### h).

With the same technique we did in g, we can conclude that we cannot reject the null hypothesis that the distribution of tme to exit from maintenance differs significantly by history of incarceration. 

```{r Q4h,echo=FALSE}

fit_exp_prison0 <- flexsurvreg(Surv(time, event) ~ 1, 
                                  data = addicts[addicts$prison ==0, ],
                                  dist = "exponential")
fit_exp_prison1 <- flexsurvreg(Surv(time, event) ~ 1, 
                                  data = addicts[addicts$prison ==1, ],
                                  dist = "exponential")


delta.p <- fit_exp_prison0$res[1] - fit_exp_prison1$res[1]
delta_se.p <- sqrt(fit_exp_prison0$res[4] ^ 2 + 
                     fit_exp_prison1$res[4] ^ 2)
T_W.p <- abs(delta.p) / delta_se.p
T_W.p

2 * pnorm(-T_W.p) 
```


## Appendix
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```